---
title: "Project 2"
author: "Keya Nandi"
date: "2025-10-31"
output: html_document
---

# **Clean and load the packages**
```{r}
rm(list=ls())
library(dplyr)
library(tidyr)
library(mFilter)
library(ggplot2)
library(zoo)
```


# **PART 1**
### Load all the data 
```{r}
setwd("D:/MS Econ/Macroeconomics")

# GDP per capita
gdp <- read.csv("D:/MS Econ/Macroeconomics/GDP per capita.csv")

# unemployment rate
unemployment <- read.csv("D:/MS Econ/Macroeconomics/Unemployment rate quarterly.csv")

# interest rate (short term)
interestrate <- read.csv("D:/MS Econ/Macroeconomics/Short term interest rate.csv")

# cpi inflation rate
inflation <- read.csv("D:/MS Econ/Macroeconomics/CPI inflation rate.csv")
```


### Mutated tables and cleaned datasets
```{r}
#clean gdp
# Inspect what the base column looks like
gdp <- gdp %>%
  filter(REF_YEAR_PRICE == "2020")

# Filter and clean GDP data
gdp <- gdp %>%
  filter(Reference.area %in% c("United States", "Australia", "Germany") &
         as.numeric(substr(TIME_PERIOD, 1, 4)) >= 2010 &
         as.numeric(substr(TIME_PERIOD, 1, 4)) <= 2023) %>%
  rename(Country = Reference.area, Time = TIME_PERIOD, GDP = OBS_VALUE) %>%
  select(Country, Time, GDP)


# Filter Unemployment data
unemployment <- unemployment %>%
  filter(Reference.area %in% c("United States", "Australia", "Germany") &
         as.numeric(substr(TIME_PERIOD, 1, 4)) >= 2010 &
         as.numeric(substr(TIME_PERIOD, 1, 4)) <= 2023) %>%
  rename(Country = Reference.area, Time = TIME_PERIOD, Unemployment = OBS_VALUE) %>%
  select(Country, Time, Unemployment)

# Filter Interest Rate data
interestrate <- interestrate %>%
  filter(Reference.area %in% c("United States", "Australia", "Germany") &
         as.numeric(substr(TIME_PERIOD, 1, 4)) >= 2010 &
         as.numeric(substr(TIME_PERIOD, 1, 4)) <= 2023) %>%
  rename(Country = Reference.area, Time = TIME_PERIOD, Interestrate = OBS_VALUE) %>%
  select(Country, Time, Interestrate)

# Filter Inflation data
inflation <- inflation %>%
  filter(Reference.area %in% c("United States", "Australia", "Germany") &
         as.numeric(substr(TIME_PERIOD, 1, 4)) >= 2010 &
         as.numeric(substr(TIME_PERIOD, 1, 4)) <= 2023) %>%
  rename(Country = Reference.area, Time = TIME_PERIOD, Inflation = OBS_VALUE) %>%
  select(Country, Time, Inflation)


```


### Combine datasets into one panel
```{r}
combined <- gdp %>%
  left_join(unemployment, by = c("Country", "Time")) %>%
  left_join(interestrate, by = c("Country", "Time")) %>%
  left_join(inflation, by = c("Country", "Time"))

colSums(is.na(combined))

```




### Detrending with HP filter
```{r}

# Apply HP filter by country for each variable
hp_results <- combined %>%
  group_by(Country) %>%
  mutate(
    GDP_cycle = hpfilter(GDP, freq = 1600)$cycle,
    Unemployment_cycle = hpfilter(Unemployment, freq = 1600)$cycle,
    Interestrate_cycle = hpfilter(Interestrate, freq = 1600)$cycle,
    Inflation_cycle = hpfilter(Inflation, freq = 1600)$cycle
  )

```

### Volatility measures
```{r}
volatility <- hp_results %>%
  group_by(Country) %>%
  summarise(
    GDP_vol = sd(GDP_cycle, na.rm = TRUE),
    Unemployment_vol = sd(Unemployment_cycle, na.rm = TRUE),
    Interest_vol = sd(Interestrate_cycle, na.rm = TRUE),
    Inflation_vol = sd(Inflation_cycle, na.rm = TRUE)
  )

volatility


```

### Correlation with GDP
```{r}
correlations <- hp_results %>%
  group_by(Country) %>%
  summarise(
    Corr_Unemployment = cor(GDP_cycle, Unemployment_cycle, use = "complete.obs"),
    Corr_Interest = cor(GDP_cycle, Interestrate_cycle, use = "complete.obs"),
    Corr_Inflation = cor(GDP_cycle, Inflation_cycle, use = "complete.obs")
  )

correlations


```

### Visualize
```{r}
library(zoo)  # for yearqtr

hp_results <- hp_results %>%
  mutate(Time = as.yearqtr(Time, format = "%Y-Q%q"))


library(ggplot2)

ggplot(hp_results, aes(x = Time, y = GDP_cycle, color = Country, group = Country)) +
  geom_line(size = 1) +
  labs(
    title = "Cyclical Component of Real GDP (HP Filtered)",
    x = "Year",
    y = "GDP Cycle"
  ) +
  scale_x_yearqtr(format = "%Y", n = 7) +  # show one tick roughly every year
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


```

### plotting multiple cycles 
```{r}
hp_long <- hp_results %>%
  select(Country, Time, GDP_cycle, Unemployment_cycle, Interestrate_cycle, Inflation_cycle) %>%
  pivot_longer(
    cols = ends_with("_cycle"),
    names_to = "Variable",
    values_to = "Cycle"
  )

ggplot(hp_long, aes(x = Time, y = Cycle, color = Country)) +
  geom_line(size = 1) +
  facet_wrap(~ Variable, scales = "free_y", ncol = 2) +
  scale_x_yearqtr(format = "%Y", n = 7) +
  labs(title = "Business Cycle Fluctuations (HP Filtered)", x = "Year", y = "Cyclical Component") +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

# **Part B**
### Load all the data 
```{r}
rm(list=ls())
setwd("D:/MS Econ/Macroeconomics")

# GDP per capita
gdp_usa <- read.csv("D:/MS Econ/Macroeconomics/GDP per capita.csv")

# unemployment rate
unemployment_usa <- read.csv("D:/MS Econ/Macroeconomics/Unemployment rate quarterly.csv")

# interest rate (short term)
interestrate_usa <- read.csv("D:/MS Econ/Macroeconomics/Short term interest rate.csv")

# cpi inflation rate
inflation_usa <- read.csv("D:/MS Econ/Macroeconomics/CPI inflation rate.csv")


```


### Mutated tables and cleaned datasets
```{r}
#clean gdp
# Inspect what the base column looks like
gdp_usa <- gdp_usa %>%
  filter(REF_YEAR_PRICE == "2020")

# Filter and clean GDP data
gdp_usa <- gdp_usa %>%
  filter(Reference.area %in% c("United States") &
         as.numeric(substr(TIME_PERIOD, 1, 4)) >= 1980 &
         as.numeric(substr(TIME_PERIOD, 1, 4)) <= 2023) %>%
  rename(Country = Reference.area, Time = TIME_PERIOD, GDP = OBS_VALUE) %>%
  select(Country, Time, GDP)


# Filter Unemployment data
unemployment_usa <- unemployment_usa %>%
  filter(Reference.area %in% c("United States") &
         as.numeric(substr(TIME_PERIOD, 1, 4)) >= 1980 &
         as.numeric(substr(TIME_PERIOD, 1, 4)) <= 2023) %>%
  rename(Country = Reference.area, Time = TIME_PERIOD, Unemployment = OBS_VALUE) %>%
  select(Country, Time, Unemployment)

# Filter Interest Rate data
interestrate_usa <- interestrate_usa %>%
  filter(Reference.area %in% c("United States") &
         as.numeric(substr(TIME_PERIOD, 1, 4)) >= 1980 &
         as.numeric(substr(TIME_PERIOD, 1, 4)) <= 2023) %>%
  rename(Country = Reference.area, Time = TIME_PERIOD, Interestrate = OBS_VALUE) %>%
  select(Country, Time, Interestrate)

# Filter Inflation data
inflation_usa <- inflation_usa %>%
  filter(Reference.area %in% c("United States") &
         as.numeric(substr(TIME_PERIOD, 1, 4)) >= 1980 &
         as.numeric(substr(TIME_PERIOD, 1, 4)) <= 2023) %>%
  rename(Country = Reference.area, Time = TIME_PERIOD, Inflation = OBS_VALUE) %>%
  select(Country, Time, Inflation)


```

### Combine datasets into one panel
```{r}
library(dplyr)

combined_usa <- gdp_usa %>%
  left_join(unemployment_usa, by = c("Country", "Time")) %>%
  left_join(interestrate_usa, by = c("Country", "Time")) %>%
  left_join(inflation_usa, by = c("Country", "Time"))

colSums(is.na(combined_usa))

```

### Ensure numeric columns
```{r}
combined_usa <- combined_usa %>%
  mutate(YearQ = as.yearqtr(Time, format = "%Y-Q%q"))

# ensure numeric columns
combined_usa <- combined_usa %>%
  mutate(
    GDP = as.numeric(GDP),
    Unemployment = as.numeric(Unemployment),
    Interestrate = as.numeric(Interestrate),
    Inflation = as.numeric(Inflation)
  ) %>%
  arrange(YearQ)

colSums(is.na(combined_usa))
```




### Detrend
```{r}

# Apply HP filter for GDP
combined_usa <- combined_usa %>%
  mutate(
    GDP_cycle = hpfilter(GDP, freq = 1600)$cycle,       # quarterly frequency
    Unemployment_cycle = hpfilter(Unemployment, freq = 1600)$cycle,
    Interestrate_cycle = hpfilter(Interestrate, freq = 1600)$cycle,
    Inflation_cycle = hpfilter(Inflation, freq = 1600)$cycle
  )
```


### Define crisis window
```{r}

gfc_start <- as.yearqtr("2007 Q4")
gfc_end   <- as.yearqtr("2009 Q2")

covid_start <- as.yearqtr("2020 Q1")
covid_end   <- as.yearqtr("2021 Q2")

```

### Compute peak-to-trough declines and recovery
```{r}
# Subset GFC period
gfc_data <- combined_usa %>% filter(YearQ >= gfc_start & YearQ <= gfc_end)

# GDP peak and trough
gdp_peak <- max(gfc_data$GDP)
gdp_trough <- min(gfc_data$GDP)
gdp_decline <- (gdp_trough - gdp_peak) / gdp_peak * 100  # % decline

```

### Visualize
```{r}
library(ggplot2)

ggplot(combined_usa, aes(x = YearQ)) +
  geom_line(aes(y = GDP_cycle, color = "GDP")) +
  geom_line(aes(y = Unemployment_cycle, color = "Unemployment")) +
  geom_line(aes(y = Inflation_cycle, color = "Inflation")) +
  geom_vline(xintercept = as.numeric(gfc_start), linetype="dashed") +
  geom_vline(xintercept = as.numeric(gfc_end), linetype="dashed") +
  geom_vline(xintercept = as.numeric(covid_start), linetype="dashed") +
  geom_vline(xintercept = as.numeric(covid_end), linetype="dashed") +
  labs(title = "US Macroeconomic Cycles: GFC", y="Deviation from trend") +
  theme_minimal()

```

### Correlation and co-movememnt
```{r}
combined_usa %>%
  summarise(
    Corr_GDP_Unemployment = cor(GDP_cycle, Unemployment_cycle, use = "complete.obs"),
    Corr_GDP_Inflation    = cor(GDP_cycle, Inflation_cycle, use = "complete.obs"),
    Corr_GDP_Interest     = cor(GDP_cycle, Interestrate_cycle, use = "complete.obs")
  )

```


# **Part C**

Load the dataset
```{r}

# 0. load
rm(list=ls())
compustat <- read.csv("D:/MS Econ/Macroeconomics/Compustat data.csv", stringsAsFactors = FALSE)

library(dplyr)
library(ggplot2)
library(zoo) # for quarters


# Keep only relevant variables and convert to numeric
compustat <- compustat %>%
  select(gvkey, datadate, datafqtr, sic, saleq, capxy, dlcq, dlttq, ceqq, niq, atq) %>%
  mutate(
    saleq = as.numeric(saleq),
    capxy = as.numeric(capxy),
    dlcq = as.numeric(dlcq),
    dlttq = as.numeric(dlttq),
    ceqq = as.numeric(ceqq),
    niq = as.numeric(niq),
    atq = as.numeric(atq)
  ) %>%
  # Remove rows where all numeric columns are NA
  filter(!(is.na(saleq) & is.na(capxy) & is.na(dlcq) & is.na(dlttq) & is.na(ceqq) & is.na(niq) & is.na(atq))) %>%
  # Fill remaining NAs with 0 for calculations
  mutate(across(c(saleq, capxy, dlcq, dlttq, ceqq, niq, atq), ~ifelse(is.na(.), 0, .)))

colSums(is.na(compustat))

#constructing firm panel
# Assuming your raw Compustat dataset is 'compustat'
compustat_clean <- compustat %>%
  arrange(gvkey, datafqtr) %>%  # make sure data is sorted by firm and time
  group_by(gvkey) %>%
  mutate(
    # Revenue growth: relative to previous quarter
    RevenueGrowth = ifelse(lag(saleq) > 0, (saleq - lag(saleq)) / lag(saleq), NA),
    
    # Investment ratio: capital expenditure relative to previous quarter's total assets
    InvestmentRatio = ifelse(lag(atq) > 0, capxy / lag(atq), NA),
    
    # Leverage: total debt / equity (if equity > 0)
    Leverage = ifelse(ceqq > 0, (dlcq + dlttq) / ceqq, NA),
    
    # Profitability ratios: ROA = net income / total assets, ROE = net income / equity
    ROA = ifelse(atq > 0, niq / atq, NA),
    ROE = ifelse(ceqq > 0, niq / ceqq, NA)
  ) %>%
  ungroup()

colSums(is.na(compustat_clean))

library(dplyr)

compustat_clean <- compustat %>%
  mutate(
    RevenueProxy = ifelse(atq > 0, saleq / atq, NA),
    InvestmentRatio = ifelse(atq > 0, capxy / atq, NA),
    Leverage = ifelse(ceqq > 0, (dlcq + dlttq) / ceqq, NA),
    ROA = ifelse(atq > 0, niq / atq, NA),
    ROE = ifelse(ceqq > 0, niq / ceqq, NA)
  )

compustat_clean <- compustat %>%
  filter(!is.na(atq) & atq != 0 & !is.na(ceqq) & ceqq != 0) %>%
  mutate(
    RevenueProxy = saleq / atq,
    InvestmentRatio = capxy / atq,
    Leverage = (dlcq + dlttq) / ceqq,
    ROA = niq / atq,
    ROE = niq / ceqq
  )


compustat_clean <- compustat_clean %>%
  arrange(gvkey, datafqtr) %>%
  mutate(
    # Extract year and quarter
    year = as.numeric(substr(datafqtr, 1, 4)),
    quarter = substr(datafqtr, 6, 7)
  )


#define periods
compustat_clean <- compustat_clean %>%
  mutate(
    Period = case_when(
      year < 2007 ~ "Pre-GFC",        # all years before 2007
      year >= 2007 & year <= 2009 ~ "GFC",  # 2007–2009
      year >= 2010 ~ "Post-GFC",      # 2010 onwards
      TRUE ~ NA_character_
    )
  )

colSums(is.na(compustat_clean))

#sector classification
compustat_clean <- compustat_clean %>%
  mutate(
    sector = case_when(
      sic >= 1 & sic < 1000 ~ "Agriculture, Forestry & Fishing",
      sic >= 1000 & sic < 1500 ~ "Mining",
      sic >= 1500 & sic < 1800 ~ "Construction",
      sic >= 2000 & sic < 4000 ~ "Manufacturing",
      sic >= 4000 & sic < 5000 ~ "Transportation & Public Utilities",
      sic >= 5000 & sic < 5200 ~ "Wholesale Trade",
      sic >= 5200 & sic < 6000 ~ "Retail Trade",
      sic >= 6000 & sic < 6800 ~ "Finance, Insurance & Real Estate",
      sic >= 7000 & sic < 9000 ~ "Services",
      sic >= 9000 & sic < 10000 ~ "Public Administration",
      TRUE ~ "Unknown"
    )
  )


colSums(is.na(compustat_clean))

#size classification
# Compute pre-GFC average total assets per firm
# Step 1: Calculate Pre-GFC average assets per firm
preGFC_assets <- compustat_clean %>%
  filter(Period == "Pre-GFC") %>%
  group_by(gvkey) %>%
  summarise(preGFC_assets = mean(atq, na.rm = TRUE))  # avg assets Pre-GFC


# Step 2: Join back to main dataset
compustat_clean <- compustat_clean %>%
  left_join(preGFC_assets, by = "gvkey")

# Step 3: Create size groups (terciles)
compustat_clean <- compustat_clean %>%
  mutate(
    size_group = case_when(
      preGFC_assets <= quantile(preGFC_assets, 1/3, na.rm = TRUE) ~ "Small",
      preGFC_assets <= quantile(preGFC_assets, 2/3, na.rm = TRUE) ~ "Medium",
      preGFC_assets > quantile(preGFC_assets, 2/3, na.rm = TRUE) ~ "Large",
      TRUE ~ NA_character_
    )
  )


colSums(is.na(compustat_clean))

#aggregate firm metrics by period
compustat_panel <- compustat_clean %>%
  group_by(gvkey) %>%
  filter(!is.na(RevenueProxy)) %>%
  ungroup()

firm_summary <- compustat_clean %>%
  group_by(Period) %>%
  summarise(
    avg_revgrowth = mean(RevenueProxy, na.rm = TRUE),
    avg_investment = mean(InvestmentRatio, na.rm = TRUE),
    avg_leverage = mean(Leverage, na.rm = TRUE),
    avg_roa = mean(ROA, na.rm = TRUE),
    avg_roe = mean(ROE, na.rm = TRUE)
  )

print(firm_summary)


```


### Time Series plot with GFC shaded
```{r}
# Compute quarterly averages for time series
ts_data <- compustat_clean %>%
  group_by(year, quarter) %>%
  summarise(
    avg_revgrowth = mean(RevenueProxy, na.rm = TRUE),
    avg_roa = mean(ROA, na.rm = TRUE)
  ) %>%
  mutate(YearQ = paste0(year, "-Q", quarter))

ggplot(ts_data, aes(x = YearQ, y = avg_revgrowth, group = 1)) +
  geom_line(color = "blue") +
  geom_vline(xintercept = which(ts_data$YearQ == "2007-Q4"), linetype = "dashed", color = "red") +
  geom_vline(xintercept = which(ts_data$YearQ == "2009-Q4"), linetype = "dashed", color = "red") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Average Revenue Growth Over Time",
       subtitle = "Red dashed lines indicate GFC period",
       x = "Quarter", y = "Revenue Growth")


```


### Sectoral/size differences
```{r}
# By sector
sector_summary <- compustat_clean %>%
  group_by(Period, sector) %>%
  summarise(avg_revgrowth = mean(RevenueProxy, na.rm = TRUE),
            avg_roa = mean(ROA, na.rm = TRUE))

# By size
size_summary <- compustat_clean %>%
  group_by(Period, size_group) %>%
  summarise(avg_revgrowth = mean(RevenueProxy, na.rm = TRUE),
            avg_roa = mean(ROA, na.rm = TRUE))


ggplot(sector_summary, aes(x = Period, y = avg_revgrowth, fill = sector)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Revenue Growth by Sector Across GFC Periods")

ggplot(size_summary, aes(x = Period, y = avg_revgrowth, fill = size_group)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Revenue Growth by Firm Size Across GFC Periods")


```





